---
title: "Using Generalized Linear Models"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Using Generalized Linear Models}
output:
  knitr:::html_vignette:
    toc: yes
---

```{r check-for-build, echo = FALSE, results='asis'}
eval_chunks <- as.logical(Sys.getenv("local_vignette_build", FALSE))
# Change this via `Sys.setenv(local_vignette_build = "TRUE")`
if(!eval_chunks) 
  cat(
    "(These documents take a long time to create, so only the code",
    "is shown here. The full version is at",
    "[https://tidymodels.github.io/embed](https://tidymodels.github.io/embed).)"
    )
```


```{r setup, include = FALSE}
library(tidymodels)
library(embed)
data(okc)
```

This method uses a generalized linear model to estimate the effect of each level of a factor predictor on the outcome. These values are retained to serve as the new encodings for the factor levels. This is sometimes referred to as _likelihood encodings_. `embed` has two estimation methods for accomplishing this: with and without pooling. 

The example used here is the OkCupid data from [Kim and Escobedo-Land (2015)(pdf)](http://www.amstat.org/publications/jse/v23n2/kim.pdf). In [Kuhn and Johnson (2018)](http://feat.engineering), these data are used to predict whether a person is in the STEM fields (science, technology, engineering, and mathematics). One predictor, geographic location, is a factor variable. The frequencies of location in the data set used here vary between `r min(table(okc$location))` person and `r max(table(okc$location))` per location. There are `r length(table(okc$location))` locations in the data. Rather than producing `r length(table(okc$location)) - 1` indicator variables for a model, a single numeric variable can be used to represent the _effect_ or _impact_ of the factor level on the outcome. In this case, where a factor outcome is being predicted (STEM or not), the effects are quantified by the log-odds of the location for being STEM.  

We first calculate the raw log-odds for the data (independent of any model):

```{r raw-data, eval = eval_chunks}
library(tidymodels)
library(embed)
data(okc)

props <- 
  okc %>%
  group_by(location) %>%
  summarise(
    prop = mean(Class == "stem"),
    log_odds  = log(prop/(1-prop)),
    n = length(Class)
  ) 
props

# later, for plotting
rng <- extendrange(props$log_odds[is.finite(props$log_odds)], f = 0.1)
```

In subsequent sections, a logistic regression model is used. When the outcome variable is numeric, the steps automatically use linear regression models to estimate effects. 

## No Pooling

In this case, the effect of each location can be estimated separately for each factor level. One method for conducting this estimation step is to fit a logistic regression with the STEM classification as the outcome and the location as the predictor. From this, the log-odds are naturally estimated by logistic regression. 

For these data, a recipe is created and `step_lencode_glm` is used:

```{r simple-glm, eval = eval_chunks}
okc_glm <- 
  recipe(Class ~ ., data = okc) %>%
  # specify the variable being encoded and the outcome
  step_lencode_glm(location, outcome = vars(Class)) %>%
  # estimate the effects
  prep(training = okc)
```

The `tidy` method can be used to extract the encodings and are merged with the raw estimates:

```{r simple-glm-extract, eval = eval_chunks}
estimates <- tidy(okc_glm, number = 1)
estimates

estimates <- 
  estimates %>% 
  dplyr::select(-terms) %>%
  set_names(c("location", "glm")) %>%
    inner_join(props, by = "location")
```

For the locations with `n > 1`, the estimates are effectively the same:

```{r simple-glm-check, eval = eval_chunks}
estimates %>%
  dplyr::filter(is.finite(log_odds)) %>%
  mutate(difference = log_odds-glm) %>%
  dplyr::select(difference) %>%
  summary()
```

Note that there is also a effect that is used for a novel location for future data sets that is the average effect:

```{r simple-glm-new, eval = eval_chunks}
tidy(okc_glm, number = 1) %>%
  dplyr::filter(level == "..new") 
```

## Partial Pooling

This method estimates the effects by using all of the locations at once using a hierarchical Bayesian generalized linear model. The locations are treated as a random set that contributes a random intercept to the previously used logistic regression. 

Partial pooling estimates each effect as a combination of the separate empirical estimates of the log-odds and the prior distribution. For locations with small sample sizes, the final estimate is _shrunken_ towards the overall mean of the log-odds. This makes sense since we have poor information for estimating these locations. For locations with many data points, the estimates reply more on the empirical estimates. [This page](https://cran.r-project.org/web/packages/rstanarm/vignettes/glmer.html) has a good discussion of pooling using Bayesian models. 
 
### Bayesian Methods

One appraoch to partial pooling is the function `step_lencode_bayes` uses the `stan_glmer` function in the `rstanarm` package. There are a number of options that can be used to control the model estimation routine, including:

```{r stan-options, eval = eval_chunks}
opts <- 
  list(
    ## the number of chains
    chains = 4,
    ## how many cores to use 
    cores = 4,
    ## the total number of iterations per chain (low here for time)
    iter = 500,
    ## set the random number seed
    seed = 8779
  )
```

Using the default priors, the model is estimated via:

```{r stan-fit-defaults, eval = eval_chunks}
okc_glmer <- 
  recipe(Class ~ ., data = okc) %>%
  step_lencode_bayes(
    location,
    outcome = vars(Class),
    options = opts
  ) %>% 
  prep(training = okc)
```

This took more time than the simple non-pooled model. The embeddings are extracted in the same way:

```{r stan-extract, eval = eval_chunks}
estimates <- 
  tidy(okc_glmer, number = 1) %>% 
  dplyr::select(-terms) %>%
  set_names(c("location", "glmer")) %>%
    inner_join(estimates, by = "location")
estimates %>% dplyr::select(location, log_odds, glm, glmer)
```

Note that the `n = 1` locations have estimates that are less extreme that the naive estimates. Also, 

Let's see the effect of the shrinkage indued by partial pooling by plotting the naive results versus the new results (finite data only):

```{r stan-compare, eval = eval_chunks}
estimates %>%
  dplyr::filter(is.finite(log_odds)) %>%
  ggplot(aes(x = log_odds, y = glmer)) + 
  geom_abline(col = "red", alpha = .5) + 
  geom_point(aes(size = sqrt(n)), alpha = .5) + 
  xlim(rng) + ylim(rng) +
  theme_bw()
```

New levels are encoded as:

```{r glmer-new, eval = eval_chunks}
tidy(okc_glmer, number = 1) %>%
  dplyr::filter(level == "..new") 
```

### Empirical Bayesian Methods/Mixed Models

The same generalized linear model can be fit using mixed models via a random intercept. The `lme4` package can also be used to get pooled estimates via `step_lencode_mixed`.

```{r mixed-rec, eval = eval_chunks}
okc_mixed <- recipe(Class ~ ., data = okc) %>%
  step_lencode_mixed(
    location,
    outcome = vars(Class),
  ) %>% 
  prep(training = okc)

estimates <- tidy(okc_mixed, number = 1) %>% 
  dplyr::select(-terms) %>%
  set_names(c("location", "mixed")) %>%
    inner_join(estimates, by = "location")
estimates %>% dplyr::select(location, log_odds, glm, glmer, mixed)
```

Comparing the raw and mixed model estimates:

```{r mixed-compare, eval = eval_chunks}
estimates %>%
  dplyr::filter(is.finite(log_odds)) %>%
  ggplot(aes(x = log_odds, y = mixed)) + 
  geom_abline(col = "red", alpha = .5) + 
  geom_point(aes(size = sqrt(n)), alpha = .5) + 
  xlim(rng) + ylim(rng) +
  theme_bw()
```

These values are very similar to the Bayesian estimates. 
