---
title: "Using Generalized Linear Models"
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Using Generalized Linear Models}
output:
  knitr:::html_vignette:
    toc: yes
---

```{r}
#| label: setup
#| include: false
library(tidymodels)
library(ggiraph)
library(embed)
library(modeldata)

data(grants)
```

This method uses a generalized linear model to estimate the effect of each level of a factor predictor on the outcome. These values are retained to serve as the new encodings for the factor levels. This is sometimes referred to as _likelihood encodings_. `embed` has two estimation methods for accomplishing this: with and without pooling. 

The example used here is the Grant data from Kuhn and Johnson (2013), these data are used to predict whether a grant application was accepted. One predictor, the sponsor, is represented by the factor variable `sponsor_code`. The frequencies of sponsors in the data set used here vary between `r min(table(grants_other$sponsor_code))` person and `r max(table(grants_other$sponsor_code))` per code. There are `r length(table(grants_other$sponsor_code))` sponsor codes in the data. Rather than producing `r length(table(grants_other$sponsor_code)) - 1` indicator variables for a model, a single numeric variable can be used to represent the _effect_ or _impact_ of the factor level on the outcome. In this case, where a factor outcome is being predicted (accepted or not), the effects are quantified by the log-odds of the sponsor code for being accepted.  

We first calculate the raw log-odds for the data (independent of any model):

```{r}
#| label: raw-data
library(tidymodels)
library(embed)

tidymodels_prefer()
theme_set(theme_bw() + theme(legend.position = "top"))

data(grants)

props <-
  grants_other |>
  group_by(sponsor_code) |>
  summarise(
    prop = mean(class == "successful"),
    log_odds = log(prop / (1 - prop)),
    n = length(class)
  ) |>
  mutate(label = paste0(gsub("_", " ", sponsor_code), " (n=", n, ")"))

props |>
  select(-label)

# later, for plotting
rng <- extendrange(props$log_odds[is.finite(props$log_odds)], f = 0.1)
```

In subsequent sections, a logistic regression model is used. When the outcome variable is numeric, the steps automatically use linear regression models to estimate effects. 

## No Pooling

In this case, the effect of each sponsor code can be estimated separately for each factor level. One method for conducting this estimation step is to fit a logistic regression with the acceptance classification as the outcome and the sponsor code as the predictor. From this, the log-odds are naturally estimated by logistic regression. 

For these data, a recipe is created and `step_lencode_glm` is used:

```{r}
#| label: simple-glm
grants_glm <-
  recipe(class ~ ., data = grants_other) |>
  # specify the variable being encoded and the outcome
  step_lencode_glm(sponsor_code, outcome = vars(class)) |>
  # estimate the effects
  prep(training = grants_other)
```

The `tidy` method can be used to extract the encodings and are merged with the raw estimates:

```{r}
#| label: simple-glm-extract
glm_estimates <-
  tidy(grants_glm, number = 1) |>
  dplyr::select(-terms, -id)
glm_estimates

glm_estimates <-
  glm_estimates |>
  set_names(c("sponsor_code", "glm")) |>
  inner_join(props, by = "sponsor_code")
```

For the sponsor codes with `n > 1`, the estimates are effectively the same:

```{r}
#| label: simple-glm-check
glm_estimates |>
  dplyr::filter(is.finite(log_odds)) |>
  mutate(difference = log_odds - glm) |>
  dplyr::select(difference) |>
  summary()
```

Note that there is also a effect that is used for a novel sponsor code for future data sets that is the average effect:

```{r}
#| label: simple-glm-new
tidy(grants_glm, number = 1) |>
  dplyr::filter(level == "..new") |>
  select(-id)
```

## Partial Pooling

This method estimates the effects by using all of the sponsor codes at once using a hierarchical Bayesian generalized linear model. The sponsor codes are treated as a random set that contributes a random intercept to the previously used logistic regression. 

Partial pooling estimates each effect as a combination of the separate empirical estimates of the log-odds and the prior distribution. For sponsor codes with small sample sizes, the final estimate is _shrunken_ towards the overall mean of the log-odds. This makes sense since we have poor information for estimating these sponsor codes. For sponsor codes with many data points, the estimates reply more on the empirical estimates. [This page](https://cran.r-project.org/web/packages/rstanarm/vignettes/glmer.html) has a good discussion of pooling using Bayesian models. 
 
```{r}
# due to Matrix problems
knitr::knit_exit()
```

### Bayesian Methods

One appraoch to partial pooling is the function `step_lencode_bayes` uses the `stan_glmer` function in the `rstanarm` package. There are a number of options that can be used to control the model estimation routine, including:

```{r}
#| label: stan-options
opts <-
  list(
    ## the number of chains
    chains = 4,
    ## how many cores to use
    cores = 4,
    ## the total number of iterations per chain (low here for time)
    iter = 1000,
    ## set the random number seed
    seed = 8779
  )
```

Using the default priors, the model is estimated via:

```{r}
#| label: stan-fit-defaults
grants_glmer <-
  recipe(class ~ ., data = grants_other) |>
  step_lencode_bayes(
    sponsor_code,
    outcome = vars(class),
    options = opts
  ) |>
  prep(training = grants_other)
```

This took more time than the simple non-pooled model. The embeddings are extracted in the same way:

```{r}
#| label: stan-extract
all_estimates <-
  tidy(grants_glmer, number = 1) |>
  dplyr::select(-terms, -id) |>
  set_names(c("sponsor_code", "glmer")) |>
  inner_join(glm_estimates, by = "sponsor_code")

all_estimates |> dplyr::select(sponsor_code, log_odds, glm, glmer)
```

Note that the `n = 1` sponsor codes have estimates that are less extreme that the naive estimates. Also, 

Let's see the effect of the shrinkage induced by partial pooling by plotting the naive results versus the new results (finite data only):

```{r}
#| label: stan-compare
#| fig-alt: "Scatter chart. log odds along the x-axis, glmer along the y-axis. a red line appears at x=y. Points appear near the line. with smaller values of log_odds appear over the line, and larger values appear over the line."
pooled_plot <-
  all_estimates |>
  dplyr::filter(is.finite(log_odds)) |>
  ggplot(aes(x = log_odds, y = glmer)) +
  geom_abline(col = "red", alpha = .5) +
  geom_point_interactive(
    aes(size = sqrt(n), tooltip = label),
    alpha = .5
  ) +
  xlim(rng) +
  ylim(rng)

# Convert the plot to a format that the html file can handle
girafe(ggobj = pooled_plot)
```

New levels are encoded as:

```{r}
#| label: glmer-new
tidy(grants_glmer, number = 1) |>
  dplyr::filter(level == "..new") |>
  dplyr::select(-terms, -id)
```

### Empirical Bayesian Methods/Mixed Models

The same generalized linear model can be fit using mixed models via a random intercept. The `lme4` package can also be used to get pooled estimates via `step_lencode_mixed`.

```{r}
#| label: mixed-rec
grants_mixed <-
  recipe(class ~ ., data = grants_other) |>
  step_lencode_mixed(
    sponsor_code,
    outcome = vars(class),
  ) |>
  prep(training = grants_other)

all_estimates <-
  tidy(grants_mixed, number = 1) |>
  dplyr::select(-terms, -id) |>
  set_names(c("sponsor_code", "mixed")) |>
  inner_join(all_estimates, by = "sponsor_code")

all_estimates |>
  dplyr::select(sponsor_code, log_odds, glm, glmer, mixed)
```

Comparing the raw and mixed model estimates:

```{r}
#| label: mixed-compare
#| fig-alt: "Scatter chart. log odds along the x-axis, glmer along the y-axis. a red line appears at x=y. Points appear near the line. with smaller values of log_odds appear over the line, and larger values appear over the line."
mixed_plot <-
  all_estimates |>
  dplyr::filter(is.finite(log_odds)) |>
  ggplot(aes(x = log_odds, y = mixed)) +
  geom_abline(col = "red", alpha = .5) +
  geom_point_interactive(
    aes(size = sqrt(n), tooltip = label),
    alpha = .5
  ) +
  xlim(rng) +
  ylim(rng)

girafe(ggobj = mixed_plot)
```

These values are very similar to the Bayesian estimates. 
